---
lecture: "Simulations-II"
format: revealjs
metadata-files: 
  - _metadata.yml
---

{{< include _titleslide.qmd >}}

## Attribution
<br><br>

*Some of the material is based on content adapted from* 

- [*Chapter 20 of R Programming for Data Science, by R. Peng*](https://bookdown.org/rdpeng/rprogdatascience/simulation.html#generating-random-numbers)

<br><br>

## Learning Objectives 
<br><br>

- Understand the mechanisms for generating and simulating data.
- Contrast empirical and theoretical distributions.
- Use simulations to approximate probability of events or distribution functions.
- Use simulations to assess theoretical properties of random variables.
- Explore the Central Limit Theorem (CLT) and Law of Large Numbers (LLN).
- Write reproducible simulation code.
- Interpret and reflect on simulation results using plots and summary statistics. 

## Last time: Simulations

Define the main steps of a simulation:

- 1. Define **the process** we want to simulate

- 2. Specify how **randomness** enters the process

- 3. Decide **what to record** from each run

- 4. **Repeat** the process many times

- 5. **Summarize** the results across runs

We used simulation to approximate two probabilities when randomness came from a physical event

#### Today, we’ll use simulations to study properties of estimators and generate randomness from sampling and using distributions!!


## Sampling from a finite population

In lecture 3, we sampled from the `wildfire` population and compared population parameters with sample estimators.

- Population parameters are unknown but fixed

- Sample estimators are random since they are based on a random sample.

#### Every time we select a new random sample we get a different estimate.

## The Wildfire Population

```{r}
library(tidyverse)
library(infer)
library(readr)

wildfire <- read_csv("data/wildfire.csv")

head(wildfire)
dim(wildfire)
```

#### We'll focus on the numerical variable `temperature`

## Population parameters for `temperature`

```{r}
wildfire |>
  summarise(
    mean_temp = mean(temperature, na.rm = TRUE),
    med_temp = median(temperature, na.rm = TRUE),
    var_temp = var(temperature, na.rm = TRUE),
    sd_temp = sd(temperature, na.rm = TRUE),
    min_temp = min(temperature, na.rm = TRUE),
    max_temp = max(temperature, na.rm = TRUE)
  )

```

## A random sample and sample estimates

```{r}
set.seed(200)

wildfire_sample <- wildfire |> rep_sample_n(size = 100)

temperature_rv <- wildfire_sample$temperature
  
c(x_bar = mean(temperature_rv, na.rm = TRUE), 
  s = sd(temperature_rv, na.rm = TRUE))  
```

## Temperature as a random variable

Every time we take a new sample, the values of `temperature` change.

> `temperature` is a random variable and randomness comes from sampling!

Any statistic of `temperature` becomes a random variable as well. 

> for example, the mean temperature ($\bar{x}$) is a random variable

#### The changes in point estimates across samples is called **sampling variability**

## Standard errors

The standard error of an estimator measures how much the sample estimate (like the mean) is expected to vary across different random samples.

In lecture 3 we said that the standard error of the sample mean is given by:  

$$SE(\bar{x}) = \frac{\sigma}{\sqrt{n}}$$
and the population parameter $\sigma$ can be estimated with the sample standard deviation $s$.

## Let's build a simulation

<br>

Let's use simulation to study the variability of the sample mean (standard error) when sampling from a finite population.

- 1. Define **the process** we want to simulate

- 2. Specify how **randomness** enters the process

- 3. Decide **what to record** from each run

- 4. **Repeat** the process many times

- 5. **Summarize** the results across runs

## iClicker 1: simulation design

Which of the following options best describes a simulation study?

- A. Drawing a single sample from the population and computing the mean of temperature in the sample

- B. Drawing many samples from the population and computing the mean of temperature in each sample 

- C. Computing the mean of temperature from the full population

- D. Generating temperature values from a Normal distribution and computing its mean

## in R

1. Take a random sample of size $n = 7000$
2. Compute the sample mean temperature
3. Repeat this 1000 times

```{r}
set.seed(200)

pop_temp_complete <- wildfire |>
 select(temperature)  |>
 drop_na()  

n <- 7000
N <- length(pop_temp_complete)

# sample_mean_reps <- pop_temp_complete |>
# ...(size = ..., reps = ...) |> 
#	group_by(...) |> 
#	summarise(temp_mean = ...(..., na.rm=TRUE))
```


#### We now have a list of 1000 sample mean temperatures!


## Summarize

We can use the list of sample estimates to study the
sampling variability.

> Let's approximate the SE computing the standard deviation of the sample estimates


```{r}
# se_simulation <- ...(sampling_dist$sample_mean)
# se_formula <- ...(pop_temp_complete$...)/...
```

#### Something seems off ....

## Sampling from a finite population

- When sampling without replacement from a finite population observations are not independent.

- The variability of the sample mean is smaller than in model-based sampling.

- This reduction in variability is captured by the finite population correction (FPC):

$$ SE(\bar{x})= \frac{\sigma}{\sqrt{n}}*\sqrt{1-\frac{n}{N}} $$

where $N$ and $n$ are the population and the sample sizes, respectively.

##

```{r eval = FALSE}
library(ggplot2)
library(tidyr)

cum_sd <- sapply(
  seq_along(sample_mean_reps$temp_mean),
  function(k) sd(sample_mean_reps$temp_mean[1:k])
)

mean_temp_df <- data.frame(
  runs = seq_along(sample_mean_reps$temp_mean),
  se_mean_temp  = cum_sd
)

mean_temp_long <- tidyr::pivot_longer(
  mean_temp_df,
  -runs,
  names_to = "event",
  values_to = "estimate"
)

fcp <- sqrt(1 - (n / N))
se_corrected_fpc <- se_formula *fcp

ref_lines <- data.frame(
  se = c(se_simulation, se_corrected_fpc, se_formula),
  method = c("Simulation SD", "FPC-corrected formula", "Uncorrected formula")
)

t_long_run <- ggplot(mean_temp_long, aes(x = runs, y = estimate, color = event)) +
  geom_line() +
  geom_hline(
    data = ref_lines,
    aes(yintercept = se, color = method),
    linetype = "dashed"
  ) +
  scale_color_manual(
    values = c(
      "Simulation SD" = "#F8766D",        
      "FPC-corrected formula" = "#00BA38", 
      "Uncorrected formula" = "#619CFF" 
    )
  ) +
  labs(
    x = "Number of simulation runs",
    y = "Estimated SE",
    color = "",
    title = "Estimated SE of mean temperature"
  )
```

## Generating values from a distribution

Another way of generating randon numbers is using  well-known probability distributions like the Normal, Poisson, and binomial. 

### in R

- `rnorm`: generate random Normal variates 
  - arguments: `n` (sample size), `mean` and `sd`

- `rpois`: generate random Poisson variates 
  - arguments: `n`, `lambda` (rate)
  
- `runif`: generate random Uniform variates 
  - arguments: `n`, `min`, `max` (interval)

- `rbinom`: generate random binomial variates 
  - arguments: `n`, `size`, `prob`

## Example

```{r}
x <- runif(10, 2, 5) 
x
summary(x)
```

#### Everytime we run this code, it generates new values

## Setting the seed

```{r}
set.seed(200)

runif(5, 2, 5) 
rnorm(2, 0, 2)
rbinom(3, 1, 0.5)

# changing the sequence

set.seed(200)

runif(5, 2, 5)
rnorm(2, 0, 2)
rpois(4,10)
rbinom(3, 1, 0.5)
```

## Simulate a simple linear regression (SLR)

- 1) Simulate a dataset from the following model:

$$y = \beta_0 + \beta_1 \times x + \varepsilon$$
where $\varepsilon \sim \mathcal{N}(0,1)$ and $x \sim \text{Unif}(1,3)$. 

In your simulation, set the population coefficients $\beta_0 = 1$ and $\beta_1 = 2$. In real data, these are unknown and you use a sample to estimate them.

- 2. Plot the data 

- 3. Use the function `lm()` to estimate the parameters of the model using least squares estimation

> Check [this application](https://setosa.io/ev/ordinary-least-squares-regression/) to recall LS estimation

## Key takeaways

:::incremental

- Simulations can be used to: 
  - approximate probabilities, 
  - study properties of estimators, 
  - approximate distributions

- Randomness can be introduced using code that: 
  - mimics physical processes
  - randomly select outcomes from a finite population
  - generates random values from a distribution
  
:::
